Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2014-07-31 improved_state_space.tex
on to the results

The DM model is an extension of the 
\citet{royle:2004biom} $N$-mixture model, which allows for inference about spatial
variation in abundance when individuals cannot be detected with
certainty. To estimate parameters associated with both abundance 
and the detection process, the original $N$-mixture model uses replicate
observations at each site. The observations are collected during sufficiently
short time intervals such that the population can safely be
assumed to be closed with respect to births, deaths, and movement. The DM
model relaxes this closure assumption and includes explicit parameters
describing population change over time.

@@

The DM model assumes that abundance in time $t$ is a function of
abundance at time $t-1$, i.e. abundance at each site evolves as a
first order Markov process. 
\citet{dail_madsen:2011} considered several models to describe the temporal dynamics;
however, in each case they modeled $N_t$ as the sum of two random variables:
$S_{i,t}$, the number of individuals surviving from $t-1$ and not
emigrating; and $G_{i,t}$ the number of new individuals entering
the population. Their most general model was
\begin{linenomath*}
\begin{equation}
\left.\begin{aligned}
S_{i,t}|N_{i,t-1} &\sim \mathrm{Bin}(N_{i,t-1}, \omega) \\
G_{i,t}|N_{i,t-1} &\sim \mathrm{Pois}(\gamma(N_{i,t-1})) \\
N_{i,t} &= S_{i,t}+G_{i,t}
\end{aligned}\right\} \quad \text{for} \; t=2,\hdots,T
\label{eq:Nt}
\end{equation}
\end{linenomath*}
where $\omega$ is the apparent survival probability and $\gamma$
is the recruitment rate (which can depend on $N_{i,t-1}$).
\citet{dail_madsen:2011} proposed three
models for $\gamma$: the constant model,
$G_{i,t} \sim \mathrm{Pois}(\gamma)$ where recruitment does not
depend on $N_{i,t-1}$, and which simulates a ``propagule rain'' of new
individuals; the autoregressive model, $G_{i,t} \sim
\mathrm{Pois}(\gamma(N_{i,t-1}))$, which
simulates geometric or density independent growth; and the
``no-trend'' model, $\gamma = (1-\omega)\Lambda$, which assumes the
population is at equilibrium. Covariates of
$\omega$ and $\gamma$ can be easily accommodated, for example
using logit- and log-linear models respectively.

@@

\section*{Model Extensions}
\label{sec:ext}
\subsection*{Population growth models}
\begin{samepage}
Partitioning population growth into survival and recruitment
is ideal in terms of providing a mechanistic description of population
dynamics, but this is not always possible using simple count data,
especially when the sites are not closed with respect to immigration
and emigration. 
When the mechanistic model is unrealistic, we suggest replacing it
with classical population growth models that have a long history in ecology.
The exponential growth model is the simplest: 
$N_{i,t} = N_{i,t-1}e^r$ where $r$ is the intrinsic
rate of increase. 
To allow for demographic stochasticity, we can regard $N_{i,t}$ as a Poisson 
random variable, simplifying Eq.~\ref{eq:Nt} to:
\begin{linenomath*}
\begin{equation}
  N_{i,t} \sim \mathrm{Pois}(\exp(r)N_{i,t-1}).
\label{eq:exp}
\end{equation}
\end{linenomath*}

Density-dependent versions of the model are also possible.  For
example:
\begin{linenomath*}
\begin{equation}
  N_{i,t} \sim \mathrm{Pois}(N_{i,t-1}\exp(r(1-N_{i,t-1}/K)))
\label{eq:rick}
\end{equation}
\end{linenomath*}
where $K$ is the stable equilibrium of the population and $r$ is
the instantaneous population growth rate at low population
densities, and both parameters are constrained to be positive. This is
a stochastic version of the Ricker-logistic population growth model
\citep{ricker:1954}. Another option is a modified Gompertz-logistic
density-dependent model \citep{hart_gotelli:2011}:
\begin{linenomath*}
\begin{equation}
N_{i,t} \sim \mathrm{Pois}(N_{i,t-1}\exp(r(1-\log(N_{i,t-1}+1)/\log(K+1))))
\label{eq:gomp}
\end{equation}
\end{linenomath*}
Here the interpretations of $r$ and $K$ are similar to those in the
Ricker-logistic model. Because a single Poisson distribution controls
the distribution of $N_{i,t}$ in each of these models, the discrete 
convolution used by \citet{dail_madsen:2011} to construct the
likelihood is not required, speeding up processing time.
\end{samepage}

@@

One limitation of the geometric-recruitment, exponential,
Ricker-logistic, and Gompertz-logistic versions of the DM models all
is that they include no mechanism for a local population to recover
after extinction. 
However, these models can be generalized 
to include both internal and external (immigration) contributions
to population growth. For example, an exponential plus immigration
model is:
\begin{linenomath*}
\begin{equation}
  N_{i,t} \sim \mathrm{Pois}(\exp(r)N_{i,t-1} + \iota)
  \label{eq:expimm2}
\end{equation}
\end{linenomath*}
where $\iota$ represents the average number of immigrants per year, and is
constrained to be
positive (this is equivalent to separate Poisson processes for
growth and immigration).  
The geometric-recruitment, Ricker-logistic,
and Gompertz-logistic models can be extended to allow for immigration in the
same way.   Deterministic versions of the Ricker-logistic + immigration and the 
Gompertz-logistic + immigration do reach stable equilibriums, but the equilibriums 
are not at $K$ and cannot be solved for analytically \citep{otto_day:2007}.  Therefore, we
refer to $K$ in these models as the semi-equilibrium abundance.

@@

The negative binomial distribution used in $N$-mixture models can have
a very long right tail, and its mean to variance ratio may not be
ideal in all cases. An alternative is to consider zero-inflated
distributions, such as the zero-inflated Poisson:
\begin{linenomath*}
\begin{equation}
N_{i1} \sim \left\{
\begin{aligned}
\mbox{Pois}(0) &\; \text{with probability} \; \psi \\
\mbox{Pois}(\Lambda) &\; \text{with probability} \; (1-\psi)
\end{aligned} \right.
\label{eq:ZIP}
\end{equation}
\end{linenomath*}
where $\psi$ represents the proportion of extra zeros. 
Mechanistically, this distribution might be appropriate when not all sites are 
within the range of a species.

Under this formulation, 
detection, abundance, and zero-inflation can be modeled separately as
functions of covariates. For example, detection of a
species might depend on wind speed,
abundance on forest type and weather, and zero-inflation upon elevation and
climate.

@@

\subsection*{Environmental and demographic stochasticity}
\begin{comment}
Even though the DM model is conceptually simple, 
it includes many random effects (the $N$'s)
making it challenging to compute the likelihood or implement MCMC
algorithms. Nonetheless, in some cases additional random effects
may be of interest. For example, in conventional state-space models,
environmental stochasticity is often modeled as $r_t \sim
\mathrm{Norm}(\mu_r, \sigma_r)$.
\end{comment}
The original DM models allows for demographic stochasticity --- variation in population growth 
rate due to the randomness of birth and death processes --- 
via the binomial and Poisson distributions for apparent survival and recruitment.  Our population
growth extensions use a single Poisson distribution for abundance. 
Conceptually, this is an appropriate method of modeling demographic stochasticity 
for semelparous organisms with one generation per sampling interval (year)
\citep{bonsall_hastings:2004,melbourne_hastings:2008},
but may actually perform better for iteroparous organisms as well (see Discussion).  
Dynamics of semelparous organisms with more than one generation per year 
(many insects) may be overdispersed compared to the Poisson.  If the number of generations per
year is known, then this can be accounted for in our models by adding extra $N$'s between each
count (see Supplement 1).

One way to model environmental stochasticity --- variation in population growth
rate due to stochastic environmental conditions --- is by explicitly
modeling dynamic parameters as functions of environmental covariates, such
as weather.  For example, in the Gompertz-logistic model (Eq.~\ref{eq:gomp}), $r$ could be a 
function of temperature ($C$) and $K$ a function of rainfall ($M$):
\begin{linenomath*}
\begin{gather}
\log(r_{i,t}) = \beta^{r}_{0} + \beta^{r}_{1}C_{i,t} \nonumber \\
\log(K_{i,t}) = \beta^{K}_{0} + \beta^{K}_{1}M_{i,t}  
\label{eq:weather}
\end{gather}
\end{linenomath*}

Another way to model environmental stochasticity is as lognormal variation in observed local population 
growth rate \citep{bjornstad:2001,bonsall_hastings:2004}.  This could be applied to both density-independent and density-dependent models.
For example, for the Ricker-logistic + immigration model:
\begin{linenomath*}
\begin{equation}
N_{i,t} \sim
\mathrm{Pois}(N_{i,t-1}\exp(\nu_{i,t} + r(1-N_{i,t-1}/K)) + \iota)
\label{eq:nuRand}
\end{equation}
\end{linenomath*}
where $\nu$ is a variable with a normal distribution, mean value of 0, and standard deviation $\sigma_\nu$.  
We suggest three variations for spatial variation in environmental stochasticity.  The first has independent environmental stochasticity 
between sites, and might be appropriate where sites are widely dispersed.  In the second, 
all sites have the same environmental conditions in a given year (regional stochasticity, $\nu_{i,t} = \nu_{t}$),
which might be appropriate where the geographic or environmental range is small \citep{hanski:1998}.  
Third, where sites can be close together but the range of sites is broad, a multivariate normal distribution 
for $\nu_{i,t}$ with distance-dependent correlation coefficients 
between sites in a year might be appropriate.

@@

@@

We tested our models against violations of the Poisson process assumption for
annual dynamics.  We ran three alternative scenarios: 1) a geometric-recruitment model with
$\omega = \gamma = 0.5$; 2) a Leslie matrix model with six age classes; and 3) 
a Gompertz-logistic model with two non-overlapping generations per year (sampling occasion).   
The first two scenarios are underdispersed compared to the Poisson and the third is overdispersed.
In the first two cases, we estimated parameters with both the exponential and geometric-recruitment
models using the \textbf{R} package \texttt{unmarked}.  We compared the bias,
root mean squared error, and coverage of the model estimates as well as their Akaike's Information
Criterion (AIC) values.
For the third scenario, we estimated parameters with the Gompertz-logistic model with 
one generation per year using both maximum likelihood and MCMC approaches, as well
as allowing for two generations per year with MCMC.  We compared the bias, root mean
squared error, and coverage of all three model estimates.  Details of the robustness simulation
study can be found in Appendix A.

@@

We fitted a series of models using maximum likelihood, 
and then a series of models using MCMC. We started by testing
three models of initial abundance (Poisson, negative binomial, and
zero-inflated Poisson) with exponential dynamics
(Eq.~\ref{eq:exp}) and no covariates.  We selected the minimum AIC model from that set to test three additional
models for p: wind, first, and wind + first. We selected the minimum
AIC model from that set to test eight additional models of dynamics:
constant-recruitment, geometric-recruitment, Ricker-logistic, Gompertz-logistic, 
the previous three with immigration, and exponential + immigration.
These models require a a finite upper bound ($N_{max}$) to integrate
over when run in a maximum likelihood
framework; we used 600.

We ran the top ranked models from the maximum likelihood
analyses in a Bayesian framework with non-informative priors ($\Lambda$,
$\alpha$, and $K  \sim \mathrm{U}(0, 200)$, $r  \sim \mathrm{U}(0, 5)$,
$\iota  \sim \mathrm{U}(0, 15)$, and logit-linear $p$ coefficients $\sim \mathrm{N}(0, 100)$).  
We added random observer 
effects and environmental stochasticity, modeled as lognormal
variation in regional population growth rate, Eq.~\ref{eq:nuRand}.
We tested for lack of convergence using
five Markov chains for each model \citep{gelman_rubin:1992}.
For each chain we sampled the MCMC for at least 40,000 iterations, after at
least 3,000 adaptation and tuning samples.  
We also used the random observer effects model to obtain route and 
year specific abundance estimates.  Because of the memory requirements involved,
we used a single chain.

>>>>

The DM model is an extension of the 
\citet{royle:2004biom} $N$-mixture model, which allows for inference about spatial
variation in abundance when individuals cannot be detected with
certainty. To estimate both abundance and detection parameters, 
%% associated with both abundance and the detection process, 
the original $N$-mixture model uses replicate
observations at each site. The observations are collected during sufficiently
short time intervals such that the population can safely be
assumed to be closed with respect to births, deaths, and movement. The DM
model relaxes this closure assumption and includes explicit parameters
describing population change over time.

@@

The DM model assumes that abundance at time $t>1$ is a function of
abundance at time $t-1$, i.e. abundance at each site evolves as a
first order Markov process. 
\citet{dail_madsen:2011} considered several models to describe the temporal dynamics;
however, in each case they modeled $N_t$ as the sum of two random variables:
$S_{i,t}$, the number of individuals surviving from $t-1$ and not
emigrating; and $G_{i,t}$ the number of new individuals entering
the population. Their most general model was
\begin{linenomath*}
\begin{equation}
\left.\begin{aligned}
S_{i,t}|N_{i,t-1} &\sim \mathrm{Bin}(N_{i,t-1}, \omega) \\
G_{i,t}|N_{i,t-1} &\sim \mathrm{Pois}(\gamma(N_{i,t-1})) \\
N_{i,t} &= S_{i,t}+G_{i,t}
\end{aligned}\right\} \quad \text{for} \; t=2,\hdots,T
\label{eq:Nt}
\end{equation}
\end{linenomath*}
where $\omega$ is the apparent survival probability and $\gamma$
is the recruitment rate, which can depend on $N_{i,t-1}$.
\citet{dail_madsen:2011} proposed three
models for $\gamma$: the constant model,
$G_{i,t} \sim \mathrm{Pois}(\gamma)$ where recruitment does not
depend on $N_{i,t-1}$, and which simulates a ``propagule rain'' of new
individuals; the autoregressive model, $G_{i,t} \sim
\mathrm{Pois}(\gamma(N_{i,t-1}))$
%% , which simulates 
of geometric or density independent growth; and the
``no-trend'' model, $\gamma = (1-\omega)\Lambda$, which assumes the
population is at equilibrium. Covariates of
$\omega$ and $\gamma$ can be easily accommodated, for example
using logit- and log-linear models respectively.

@@

\section*{Model Extensions}
\label{sec:ext}
\subsection*{Population growth models}
\begin{samepage}
Partitioning population growth into survival and recruitment
is useful %%ideal 
in terms of providing a mechanistic description of population
dynamics, but this is not always possible using simple count data,
especially when the sites are not closed with respect to immigration
and emigration. 
When the mechanistic model is unrealistic, we suggest replacing it
with classical population growth models that have a long history in ecology.
The exponential growth model is the simplest: %% Should this geometric instead of exponential?
$N_{i,t} = N_{i,t-1}e^r$ where $r$ is the intrinsic
rate of increase. 
To allow for demographic stochasticity, we can regard $N_{i,t}$ as a Poisson 
random variable, simplifying Eq.~\ref{eq:Nt} to:
\begin{linenomath*}
\begin{equation}
  N_{i,t} \sim \mathrm{Pois}(\exp(r)N_{i,t-1}).
\label{eq:exp}
\end{equation}
\end{linenomath*}

Density-dependent versions of the model are also possible.  For
example:
\begin{linenomath*}
\begin{equation}
  N_{i,t} \sim \mathrm{Pois}(N_{i,t-1}\exp(r(1-N_{i,t-1}/K)))
\label{eq:rick}
\end{equation}
\end{linenomath*}
where $K$ is the stable equilibrium of the population and $r$ is
the instantaneous population growth rate at low population
densities, and both parameters are constrained to be positive. This is
a stochastic version of the 
Ricker-logistic population growth model %% Could we just call this the "Ricker model"?
\citep{ricker:1954}. Another option is a modified Gompertz-logistic
density-dependent model \citep{hart_gotelli:2011}:
\begin{linenomath*}
\begin{equation}
N_{i,t} \sim \mathrm{Pois}(N_{i,t-1}\exp(r(1-\log(N_{i,t-1}+1)/\log(K+1))))
\label{eq:gomp}
\end{equation}
\end{linenomath*}
Here the interpretations of $r$ and $K$ are similar to those in the
Ricker-logistic model. Because a single Poisson distribution controls
the distribution of $N_{i,t}$ in each of these models, the discrete 
convolution used by \citet{dail_madsen:2011} to construct the
likelihood is not required, speeding up processing time.
\end{samepage}

@@

One limitation of the geometric-recruitment, exponential,
Ricker-logistic, and Gompertz-logistic versions of the DM models %%all
is that they include no mechanism for a local population to recover
after extinction. 
However, these models can be generalized 
to include both internal and external (immigration) contributions
to population growth. For example, an exponential plus immigration
model is:
\begin{linenomath*}
\begin{equation}
  N_{i,t} \sim \mathrm{Pois}(\exp(r)N_{i,t-1} + \iota)
  \label{eq:expimm2}
\end{equation}
\end{linenomath*}
where $\iota$ represents the average number of immigrants per year, and is
constrained to be
positive (this is equivalent to separate Poisson processes for
growth and immigration).  
The geometric-recruitment, Ricker-logistic,
and Gompertz-logistic models can be extended to allow for immigration in the
same way.   Deterministic versions of the Ricker-logistic + immigration and the 
Gompertz-logistic + immigration do reach stable equilibriums, but the equilibriums 
are not at $K$ and cannot be solved for analytically \citep{otto_day:2007}.  Therefore, we
refer to $K$ in these models as the semi-equilibrium abundance.

@@

The negative binomial distribution used in $N$-mixture models can have
a very long right tail, and its mean to variance ratio may not be
ideal in all cases. An alternative is to consider zero-inflated
distributions, such as the zero-inflated Poisson:
\begin{linenomath*}
\begin{equation}
N_{i1} \sim \left\{
\begin{aligned}
\mbox{Pois}(0) &\; \text{with probability} \; \psi \\
\mbox{Pois}(\Lambda) &\; \text{with probability} \; (1-\psi)
\end{aligned} \right.
\label{eq:ZIP}
\end{equation}
\end{linenomath*}
where $\psi$ represents the proportion of extra zeros. 
%% Mechanistically, this distribution might be appropriate when not all sites are 
%% within the range of a species.
%%
Under this formulation, 
detection, abundance, and zero-inflation can be modeled separately as
functions of covariates. For example, detection of a
species might depend on wind speed,
abundance on forest type and weather, and zero-inflation upon elevation and
climate.

@@

\subsection*{Environmental and demographic stochasticity}
\begin{comment}
Even though the DM model is conceptually simple, 
it includes many random effects (the $N$'s)
making it challenging to compute the likelihood or implement MCMC
algorithms. Nonetheless, in some cases additional random effects
may be of interest. For example, in conventional state-space models,
environmental stochasticity is often modeled as $r_t \sim
\mathrm{Norm}(\mu_r, \sigma_r)$.
\end{comment}
The original DM models allows for demographic stochasticity --- variation in population growth 
rate due to the randomness of birth and death processes --- 
via the binomial and Poisson distributions for apparent survival and recruitment.  Our population
growth extensions use a single Poisson distribution for abundance. 
% NOTE: I would save this for the discussion
% Conceptually, this is an appropriate method of modeling demographic stochasticity 
% for semelparous organisms with one generation per sampling interval (year)
% \citep{bonsall_hastings:2004,melbourne_hastings:2008},
% but may actually perform better for iteroparous organisms as well (see Discussion).  
% Dynamics of semelparous organisms with more than one generation per year 
% (many insects) may be overdispersed compared to the Poisson.  If the number of generations per
% year is known, then this can be accounted for in our models by adding extra $N$'s between each
% count (see Supplement 1).

One way to model environmental stochasticity --- variation in population growth
rate due to stochastic environmental conditions --- is by explicitly
modeling dynamic parameters as functions of environmental covariates, such
as weather.  For example, in the Gompertz-logistic model (Eq.~\ref{eq:gomp}), $r$ could be a 
function of temperature %($C$) 
and $K$ a function of rainfall. %($M$):
%% NOTE: I don't think these equations are necessary
% \begin{linenomath*}
% \begin{gather}
% \log(r_{i,t}) = \beta^{r}_{0} + \beta^{r}_{1}C_{i,t} \nonumber \\
% \log(K_{i,t}) = \beta^{K}_{0} + \beta^{K}_{1}M_{i,t}  
% \label{eq:weather}
% \end{gather}
% \end{linenomath*}
% 
Another way to model environmental stochasticity is as lognormal variation in 
the %observed local 
population growth rate \citep{bjornstad:2001,bonsall_hastings:2004}.  
This could be applied to both density-independent and density-dependent models.
For example, for the Ricker-logistic + immigration model:
\begin{linenomath*}
\begin{equation}
N_{i,t} \sim
\mathrm{Pois}(N_{i,t-1}\exp(\nu_{i,t} + r(1-N_{i,t-1}/K)) + \iota)
\label{eq:nuRand}
\end{equation}
\end{linenomath*}
where $\nu$ is a variable with a normal distribution, mean value of 0, and standard deviation $\sigma_\nu$.  
We suggest three variations for spatial variation in environmental stochasticity.  
%% These are interesting ideas, but do we actually evaluate them? If not, it might be 
%% better to save them for the discussion.
The first has independent environmental stochasticity 
between sites, and might be appropriate where sites are widely dispersed.  In the second, 
all sites have the same environmental conditions in a given year (regional stochasticity, $\nu_{i,t} = \nu_{t}$),
which might be appropriate where the geographic or environmental range is small \citep{hanski:1998}.  
Third, where sites can be close together but the range of sites is broad, a multivariate normal distribution 
for $\nu_{i,t}$ with distance-dependent correlation coefficients 
between sites in a year might be appropriate.

@@

%\subsection*{Proof of Concept Simulation Study}
\subsection*{Simulation Study to Assess Estimator Accuracy}

@@

We tested our models against violations of the Poisson process assumption for
annual dynamics.  We ran three alternative scenarios: 1) a geometric-recruitment model with
$\omega = \gamma = 0.5$; 2) a Leslie matrix model with six age classes; and 3) 
a Gompertz-logistic model with two non-overlapping generations per year (sampling occasion).   
%% Couldn't we expand our model to deal with these "mis-specifications"? Like Elise and I did in a couple of recent papers?
The first two scenarios are underdispersed compared to the Poisson and the third is overdispersed.
In the first two cases, we estimated parameters with both the exponential and geometric-recruitment
models using the \textbf{R} package \texttt{unmarked}.  We compared the bias,
root mean squared error, and coverage of the model estimates as well as their Akaike's Information
Criterion (AIC) values.
For the third scenario, we estimated parameters with the Gompertz-logistic model with 
one generation per year using both maximum likelihood and MCMC approaches, as well
as allowing for two generations per year with MCMC.  We compared the bias, root mean
squared error, and coverage of all three model estimates.  Details of the robustness simulation
study can be found in Appendix A.

@@

We fit a series of models using maximum likelihood, 
and then a series of models using MCMC. We started by testing
three models of initial abundance (Poisson, negative binomial, and
zero-inflated Poisson) with exponential dynamics
(Eq.~\ref{eq:exp}) and no covariates.  We selected the minimum AIC model from that set to test three additional
models for p: wind, first, and wind + first. We selected the minimum
AIC model from that set to test eight additional models of dynamics:
constant-recruitment, geometric-recruitment, Ricker-logistic, Gompertz-logistic, 
the previous three with immigration, and exponential + immigration.
These models require a a finite upper bound ($N_{max}$) to integrate
over when run in a maximum likelihood
framework; we used 600.

We ran the top ranked models from the maximum likelihood
analyses in a Bayesian framework with vague %non-informative 
priors ($\Lambda$,
$\alpha$, and $K  \sim \mathrm{U}(0, 200)$, $r  \sim \mathrm{U}(0, 5)$,
$\iota  \sim \mathrm{U}(0, 15)$, and logit-linear $p$ coefficients $\sim \mathrm{N}(0, 100)$).  
We added random observer 
effects and environmental stochasticity, modeled as lognormal
variation in regional population growth rate, Eq.~\ref{eq:nuRand}.
We tested for lack of convergence using
five Markov chains for each model \citep{gelman_rubin:1992}.
For each chain, we drew at least 40,000 samples, 
%the MCMC for at least 40,000 iterations, 
after at
least 3,000 adaptation iterations. %and tuning samples.  
We also used the random observer effects model to obtain route and 
year specific abundance estimates.  Because of the memory requirements involved,
we used a single chain.
